Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-14 index.qmd
section:background,structural; process:write

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might seek to answer the question, "To what extent do daily practice hours ($X$) influence students' ability to produce high-quality written texts ($Y$)?". To investigate this, the study randomly assigns students to two groups with different levels of daily practice, low $(X=1)$ and medium $(X=2)$. The quality of the written texts is evaluated using pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$. The research question is then framed as the estimand, "*On average*, which group of students produces higher-quality written texts?" This estimand is mathematically expressed as $E[Y|X=1] - E[Y|X=2]$, where $E[\cdot]$ represents the expected value.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* is the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which is the process that integrates data with an estimator [@Everitt_et_al_2010]. While various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure accurate estimates. For example, Fisher's exact test [@Fisher_1922] is an estimator with desirable properties for comparing $2\times2$ tables. The test provides accurate estimates for comparing proportions when its assumptions are satisfied. @fig-IEflow, known as the Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], illustrates the aforementioned process of transitioning from estimands to estimates.

@@

<!-- 3. But we want these estimands to be causal -->
However, many studies aim to establish causal relationships rather than merely infer associations. In the earlier example, the estimates obtained using Fisher's exact test can be interpreted as causal because the data were collected through a randomized experiment. In randomized experiments, the researcher fully controls the treatment assignment mechanism, such as assigning students to different groups. This control allows causal estimates to be derived from associational estimates, such as those provided by the Fisher statistic, since randomization effectively eliminates confounding [@Neal_2020]. *Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious association [@Everitt_et_al_2010].

<!-- 4. causal inference to the rescue -->
<!-- Nevertheless, in many cases, researchers are not able to perform experiments due to many reasons. For instance, . In this scenarios, the framework of causal inference can be an alternative to generate causal estimates. *Causal inference* is a framework focused on identifying the causes influencing a phenomenon and estimating their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which primarily focuses on summarizing data and inferring associations, causal inference provides a coherent mathematical framework for analyzing causes and counterfactuals [@Pearl_2009]. -->

<!-- 5. How does causal inference defines causal effects? what are counterfactuals? -->
<!-- Causal inference defines causal effects in terms of counterfactuals. Counterfactuals represent scenarios *contrary to fact*, where alternative *potential* outcomes resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. Returning to the previous example, if the research aim to determine the $X$ have an impact on an outcome $Y$ -->

<!-- Causal inference approach the *fundamental problem of causal inference* -->

<!-- According to @Pearl_et_al_2018, counterfactuals are the foundation of causal inference and occupy the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association. Nevertheless, despite their abstract nature, counterfactuals enable the development of a *theory of the world* that explains why specific causes have specific effects and what occurs in their absence [@Pearl_et_al_2018]. They achieve this by translating causal statements into counterfactual statements, that is, statements about "what would have happened in the world under different circumstances." -->

<!-- 6. How do we apply causal inference and counterfactuals? -->
<!-- Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs the do-operator and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010]. -->

<!-- 7. But which approach are we going to use? -->
<!-- However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016]. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. -->

@@

>>>>

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might examine the question, "To what extent do different teaching methods ($X$) influence students' conceptual understanding of algebra ($Y$)?" To investigate this, the study randomly assigns students to two groups, each using different algebra teaching method $(X=\{1,2\})$. Students' conceptual understanding of algebra is evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student possesses a higher degree of understanding among those compared. The research question is then framed as the estimand, "*On average*, is there a difference in conceptual understanding of algebra between the two groups of student?" This estimand is mathematically expressed as $E[Y| X=1] - E[Y| X=2]$, where $E[\cdot]$ denotes the expected value. An example of this approach is seen in @Jones_et_al_2019.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that maps data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refer to the process of integrating data with an estimator [@Everitt_et_al_2010]. Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, providing accurate estimates for the desired estimand when its assumptions are met.

<!-- 3. But we want these estimands to be causal -->
However, many studies aim to understand the mechanisms underlying specific data types and also seek to establish causal relationships rather than merely infer associations. In the earlier example, the associational estimates obtained using the BTL model can be interpreted as causal, because the data were collected using a randomized experiment. Randomized experiments enable the causal interpretation of associational estimates by ensuring several key properties, including common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most crucial property, however, is that randomization effectively eliminates confounding. *Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious association [@Everitt_et_al_2010]. Randomization mitigates this issue by decoupling the intervention assignment mechanism, such as assigning students to different groups, from other variables and outcomes [@Morgan_et_al_2014; @Neal_2020].

<!-- 4. causal inference to the rescue -->
Although experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], researchers often encounter constraints that limit their ability to conduct experimental studies. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these instances, causal inference offers a valuable alternative for generating causal estimates. *Causal inference* is a framework designed to identify the causes of phenomena and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, causal inference provides a rigorous mathematical framework for analyzing causes and counterfactuals [@Pearl_2009]. Moreover, the framework is beneficial for both observational and experimental studies, as it provides significant theoretical insights that enhance their design [@McElreath_2020].

<!-- 5. But how does it work? -->
Counterfactuals represent the highest level of cognitive abstraction in the ladder of causation, followed by intervention and association [@Pearl_et_al_2018], and a detailed explanation of causes, counterfactuals, and their applications is beyond the scope of this document. However, it is important to note that causal inference addresses the question, "What would have happened in the world under different circumstances?" by transforming unobserved *potential outcomes*, known as counterfactuals, into associational estimates. *Counterfactuals* represent scenarios *contrary to fact*, where alternative *potential outcomes* resulting from a cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The framework achieves this transformation while ensuring the absence of confounding. Ultimately, the *fundamental problem of causality* is reduced to calculating differences between averages of observed outcomes and the transformed counterfactuals, thereby identifying causal effects.

<!-- 6. How do we apply causal inference and counterfactuals? -->
Several approaches to causal inference and counterfactuals exist, but two are particularly prominent: the potential outcomes approach, also known as the Neyman-Rubin causal model [@Neyman_et_al_1923; @Rubin_1974], and the structural approach [@Pearl_2009; @Pearl_et_al_2016]. Both approaches employ rigorous mathematical notation to characterize causal inference, but they do so in different ways [@Neal_2020]. The potential outcomes approach relies on counterfactual notation, whereas the structural approach employs do-calculus [@Pearl_2009]. Despite these differences, both notations can be expressed in terms of the other, and both approaches provide methods for using experimental and observational data to estimate causal effects [@Pearl_2010].

<!-- 7. But which approach are we going to use? -->
However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical and formal representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020] and and structural causal models [SCM, @Pearl_2009; @Pearl_et_al_2016]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016]. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, and the application of the sicentific model and data to identify and estimates causal effects.

@@

@@

<!-- The main assumption we’ve seen that relates the graph to the distribution -->
<!-- is the Markov assumption. The Markov assumption tells us if variables are -->
<!-- d-separated in the graph 𝐺 , then they are independent in the distribution -->
<!-- 𝑃 (Theorem 3.1): -->
<!-- Maybe we can detect independencies in the data and then use that -->
<!-- to infer the causal graph. However, going from independencies in the -->
<!-- distribution 𝑃 to d-separations in the graph 𝐺 isn’t something that the -->
<!-- Markov assumption gives us (see Equation 3.20 above). Rather, we need -->
<!-- the converse of the Markov assumption. This is known as the faithfulness -->
<!-- assumption. -->
<!-- Assumption 11.1 (Faithfulness) -->
